---
title: "Case Study_1_Cyclistic"
author: "Miguel Aldana"
date: "2024-10-30"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scenario

You are a junior data analyst working on the marketing analyst team at
Cyclistic, a bike-share company in Chicago. The director of marketing
believes the company’s future success depends on maximizing the number
of annual memberships. Therefore, your team wants to understand how
casual riders and annual members use Cyclistic bikes differently. From
these insights, your team will design a new marketing strategy to
convert casual riders into annual members. But first, Cyclistic
executives must approve your recommendations, so they must be backed up
with compelling data insights and professional data visualizations.

## Ask

Three questions will guide the future marketing program:

1\. How do annual members and casual riders use Cyclistic bikes
differently?

2\. Why would casual riders buy Cyclistic annual memberships?

3\. How can Cyclistic use digital media to influence casual riders to
become members?

#### Guiding Questions

**What is the problem you are trying to solve?**

How to convert casual riders into annual members.

**How can your insights drive business decisions?**

Understanding the differences in bike usage between casual riders and
annual members. Such as patterns, preferences, trends, customer
segmentation and service improvement.

#### Key Tasks

**Identify the business task**

Understand how annual members and casual riders use Cyclistic bikes
differently to improve our marketing strategy aimed to convert casual
riders into annual members.

**Consider key stakeholders**

1\. Cyclistic marketing analytics team

2\. Lily Moreno

3\. Cyclistic executive team

### Deliverable 1✔ {style="Color: purple"}

**Clear Statement of the Business Task**

Analyze how Cyclistic's casual riders and annual members differ in their
usage of the bike-share service.This analysis will identify patterns,
preferences and trends to improve our marketing strategy aimed at
converting casual riders into annual members. All this to increase
Cyclistic's long-term revenue and customer retention.

## Prepare

[Download the previous 12
months](https://divvy-tripdata.s3.amazonaws.com/index.html). *(In this
case we use 12 months, 202310 to 202409 "divvy-tripdata")*

Here the [License](https://divvybikes.com/data-license-agreement).

#### Guiding Questions

**Where is your data located?**

The data is located in [Download Divvy trip history
data](https://divvy-tripdata.s3.amazonaws.com/index.html), which is
publicly available and provided by Motivate International Inc.

**How is the data organized?**

The data is organized as monthly .CSV files, each file contains
information about individual trips done by Cyclistic's users, structured
data that contains 13 columns.

**Are there issues with bias or credibility in this data? Does your data
ROCCC?**

**Reliable:** Since the data is provided by the actual service operator
(Motivate International Inc.) we can say that it's credible and
accurate.

**Original:** The data is originally collected from the company system.

**Comprehensive:** The data provided cover essential data about trips
details that will be helpful with the analysis.

**Current:** The data is updated monthly so is relevant and not
out-dated.

**Cited:** The data is open source with the requirement to cite Motivate
International Inc. who is the data provider.

**How are you addressing licensing, privacy, security, and
accessibility?**

The data falls under [Divvy’s public
license](https://divvybikes.com/data-license-agreement). The data has
been anonymized and doesn't contain personal info or payments so there's
not security risk at all, accessibility is via .CSV file format.

**How did you verify the data’s integrity?**

I verified data's integrity by checking consistency in column headers
and data type in each column, outliers or missing values were identified
and will be cleaned in the process stage.

**How does it help you answer your question?**

This data helps us to analyse differences about casual riders and annual
members, such as bike types, traveled distance and trip duration.

**Are there any problems with the data?**

Missing values, wrong data types and maybe some outliers that need to be
cleaned.

#### Key Tasks

**Download data and store it appropriately**✔

```{r}
library(dplyr)
library(tidyverse)
# Set the path to the directory containing your CSV files
setwd("C:/Users/Usuario/Desktop/Google Data Analyst/8) Capstone Complete a Case Study/Case_Study_1")
# List all CSV files in the directory
file_list <- list.files(pattern = "*.csv")

# Read and bind all files into one data frame
combined_data <- file_list %>%
  lapply(read.csv) %>%    # Read each file
  bind_rows()             # Combine them into a single data frame
```

**Identify how it’s organized**✔

```{r}
# View the first few rows of the combined data
head(combined_data)
View(combined_data)
```

**Sort and filter the data**✔

**Step 1 - Fixing data type and calculate trip duration before sorting**

```{r Step 1}
library(lubridate)
glimpse(combined_data)
#1)_With glimpse we discovered that our started_at and ended_at data type is incorrect so we fix before sorting and filtering
combined_data$started_at <- as.POSIXct(combined_data$started_at, format = "%Y-%m-%d %H:%M:%S")
combined_data$ended_at <- as.POSIXct(combined_data$ended_at, format = "%Y-%m-%d %H:%M:%S")

#Calculate trip time in minutes
combined_data$trip_duration <- as.numeric(difftime(combined_data$ended_at, combined_data$started_at, units = "mins"))
```

**Step 2 - Sorting data**

```{r}
#2)_We sort our data arranging by our started_at column
combined_data <- arrange(combined_data,started_at)
```

**Step 3 - Filtering data**

```{r}
#3)_We filter rows with start_station_id and end_station_id with missing values
#3.1)_We noticed that we are not filtering properly because maybe the cells are "", not only na.
#count rows with empty values or blank spaces to ensure 3.1)
sum(combined_data$start_station_id == "" | combined_data$start_station_id == " ") 
sum(combined_data$end_station_id == "" | combined_data$end_station_id == " ") 
#Knowing that we have cells with empty values and blank spaces, we need to add this into the filter
combined_data_sort_filter <-  combined_data %>% 
  filter(!is.na(start_station_id) & start_station_id != "" & start_station_id != " ",
         !is.na(end_station_id) & end_station_id != "" & end_station_id != " ")
#Count if there're not any missing values left
sum(combined_data_sort_filter$start_station_id == "" | combined_data_sort_filter$start_station_id == " ") 
sum(combined_data_sort_filter$end_station_id == "" | combined_data_sort_filter$end_station_id == " ")

#Now that we know that everything is ok, we can continue with our original name "combined_data"
combined_data <- combined_data_sort_filter
#With rm we remove the data frame that we are not using anymore
rm(combined_data_sort_filter)
```

**Determine the credibility of the data**✔

### Deliverable 2✔ {style="Color: purple"}

**A description of all data sources used**

The description of all data sources used are on the guiding questions of
this step and the R Chunk.

## Process

#### Guiding Questions

**What tools are you choosing and why?**

I am using R because we are using a large amount of data and this
software let us work perfectly with it. Also, R provides packages like
*dplyr*,*tidyr* and *ggplot2* for efficient data manipulation,
visualization and analysis. Additionally, RStudio makes it easy to
create reports and document my entire process with R Markdown.

**Have you ensured the integrity of your data?**

Yes, we did it on the **Prepare** step such as:

-Verified that all datasets from the 12 months were successfully loaded
and combined without any loss of information.

-Checked inconsistencies between columns like start_station_id and
end_station_id.

-Ensured proper handling of missing values and removed rows where
critical fields (e.g., station IDs) were empty.

**What steps have you taken to ensure your data is clean?**

The following cleaning steps were performed:

-Converted date and time columns from character to datetime formats to
enable proper sorting and analysis.

-Removed rows with NA values in essential columns like start_station_id
and end_station_id.

-Filtered out erroneous data, such as trips with negative durations or
outlier distances.

**How can you verify that your data is clean and ready to analyze?**

**-Missing values** addressed.

**-Correct data types** for all columns.

**-No duplicates** or invalid records.

**-Logical consistency** in relationships.

**-Consistent categories** for factors.

**Have you documented your cleaning process so you can review and share
those results?**

Yes i did, i use notes on each parts of the code to document it.

#### Key Tasks

**Check your data for errors.**✔

**Choose your tools.**✔

**Transform your data so you can work with it effectively.**✔

**Document the cleaning process.**✔

### Deliverable 3✔ {style="Color: purple"}

**Documentation of any data cleansing or manipulation**

We can see the documentation of the cleaning and data manipulation in
the R Chunk.

```{r}
#Check for Missing Values
# Summary of missing values
colSums(is.na(combined_data))

#Check Data Types
# View data types
str(combined_data)

#Check for Duplicates
# Count duplicates
sum(duplicated(combined_data))
# Remove duplicates
combined_data <- combined_data[!duplicated(combined_data), ]
#Count duplicates again to double check
sum(duplicated(combined_data))

#Check for Outliers and verify logical consistency
# Summary of numerical columns
summary(combined_data)
# Remove Abnormal Trip_duration (Negative and more than 1440)
combined_data <- combined_data %>% filter(trip_duration > 1 & trip_duration < 1440)
# Summary again
summary(combined_data)

#Check for Inconsistent Categories (Typo)
# Unique values in categorical columns
unique(combined_data$rideable_type)
unique(combined_data$member_casual)
```

## Analyze

#### Guiding Questions

**How should you organize your data to perform an analysis?**

The data should be organized by key variables that allow for meaningful
comparisons and aggregations.

**Is your data properly formatted?**

Yes, the data has been cleaned and formatted for analysis on previous
steps and here.

**What surprises did you discover in the data?**

Similar bike preferences, Weekend peaks for casual riders, Trip duration
contrast and that we only have unique id's in the ride_id column.

**What trends or relationships did you find in the data?**

-Trip duration: Casual riders have significantly longer trips than
members, possibly due to leisure-oriented use.

-Weekly patterns: Members use bikes more on weekdays, aligning with
commuting purposes, while casual riders dominate on weekends.

-Consistent distances: Both groups travel similar distances per trip,
regardless of rider type. These relationships help segment user behavior
and identify key areas for targeting marketing efforts.

**How will these insights help you answer your business questions?**

-Understanding usage patterns: By identifying when and how each group
uses the service, marketing strategies can be tailored to their needs.

-Improving member acquisition: Insights into casual riders' longer trips
and weekend usage highlight opportunities to incentive membership
through leisure-focused campaigns.

-Enhancing customer retention: Knowing members' preference for weekday
commuting can guide improvements in services and promotions aimed at
enhancing their daily experience.

#### Key Tasks

**Aggregate your data to make it useful and accessible.**✔

**Organize and format your data.**✔

**Perform calculations.**✔

**Identify trends and relationships.**✔

```{r Install Needed Packages}
install.packages("geosphere")
library(geosphere)
library(ggplot2)
```

```{r Column distance}
#Adding new column with distance traveled
combined_data$distance_km <- distHaversine(
  matrix(c(combined_data$start_lng,combined_data$start_lat),ncol = 2),
  matrix(c(combined_data$end_lng,combined_data$end_lat),ncol = 2)
)/1000 #Converts m to km
```

```{r Grouping 1}
#Aggregate data by rideable_type
aggregated_data_rideable <- combined_data %>%
  group_by(rideable_type) %>%
  summarise(
    avg_duration_mins = mean(trip_duration, na.rm = TRUE),
    avg_distance_km = mean(distance_km, na.rm = TRUE),
    total_rides = n()
  )
print(aggregated_data_rideable)
```

```{r Grouping 2}
#Aggregate data by member_casual
aggregated_data_member_casual <- combined_data %>%
  group_by(member_casual) %>%
  summarise(
    avg_duration_mins = mean(trip_duration, na.rm = TRUE),
    avg_distance_km = mean(distance_km, na.rm = TRUE),
    total_rides = n(),
    classic_bike = sum(rideable_type == "classic_bike", na.rm = TRUE),
    electric_bike = sum(rideable_type == "electric_bike", na.rm = TRUE),
    electric_scooter = sum(rideable_type == "electric_scooter", na.rm = TRUE)
     )
print(aggregated_data_member_casual)
```

```{r Tranform to Graph 1}
# Transform to a long format to viz the graph easier
aggregated_long <- aggregated_data_member_casual %>%
  pivot_longer(cols = c(classic_bike, electric_bike, electric_scooter), 
               names_to = "bike_type", 
               values_to = "count")
```

```{r Bar Graph Rider Type, Bike type and Number of rides (Graph 1)}
# Create a bar graph
ggplot(aggregated_long, aes(x = member_casual, y = count, fill = bike_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Comparison of Bike Usage by Member and Casual Riders (Graph 1)",
    x = "Rider Type",
    y = "Number of Rides",
    fill = "Bike Type"
  ) +
  theme_minimal()
```

```{r Tranform to Graph 2}
# Create a bar graph for a better look of average trip duration and distance between casual - member
aggregated_data_duration_distance_long <- aggregated_data_member_casual %>%
  pivot_longer(cols = c(avg_duration_mins, avg_distance_km), 
               names_to = "metric", 
               values_to = "value")
```

```{r Bar Graph Distance, Duration and Rider Type (Graph 2)}
ggplot(aggregated_data_duration_distance_long, aes(x = member_casual, y = value, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +  # Create the bar for both metrics
  facet_wrap(~metric, scales = "free_y") +  # Facet metrics
  labs(
    title = "Average Trip Duration and Distance by Member Type (Graph 2)",
    x = "Rider Type",
    y = "Values"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("casual" = "blue", "member" = "green"))
```

```{r Grouping 3}
# Group by day of the week and type of user
daily_usage <- combined_data %>%
  mutate(day_of_week = weekdays(started_at)) %>%
  group_by(member_casual, day_of_week) %>%
  summarise(total_rides = n(), .groups = "drop")
```

```{r Line graph day of the week and rider type (Graph 3)}
# Create a line graph to compare
ggplot(daily_usage, aes(x = day_of_week, y = total_rides, color = member_casual, group = member_casual)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Cyclistic Bike Usage by Day of the Week (Graph 3)",
    x = "Day of the Week",
    y = "Number of Rides",
    color = "Rider Type"
  ) +
  theme_minimal()
```

### Deliverable 4✔ {style="Color: purple"}

**A summary of my analysis**

Initially, we can observe that members outnumber casual riders by 44.7%.
To make meaningful comparisons between the rider types, it is essential
to analyze the duration of each trip, the distance traveled, and the
type of bike used. From this analysis (as seen in Graph 1 and 2), we
found that both types of customers are quite similar. However, a
significant difference emerged in the duration of their trips. Although
both groups travel similar distances on average and prefer the same type
of bike, members' trips are almost 50% shorter in duration. This could
be because members use the bikes more regularly, while casual riders do
not.

To delve deeper into this hypothesis, we analyzed bike usage by day of
the week (Graph 3). This confirmed that members use the bikes more
frequently than casual riders. Casual riders exhibit higher usage during
weekends, whereas members primarily use the bikes on weekdays. This
suggests that members may use the bikes for commuting to work or school,
while casual riders tend to use them for leisure activities on weekends.

## Share

#### Guiding Questions

**Were you able to answer the question of how annual members and casual
riders use cycling bikes differently?**

Yes, the analysis provided clear insights into how annual members and
casual riders differ in their usage patterns:

✔ Duration ✔ Timing ✔ Purpose

**What story does your data tell?**

The data tells the story of two distinct customer segments:

Members: Efficient and frequent users who rely on bikes for short,
consistent trips during weekdays, likely for commuting purposes.

Casual Riders: Sporadic users who prefer longer trips on weekends,
indicating a recreational use of the service. This dual behavior
reflects different needs and motivations for using the service, guiding
Cyclistic’s marketing and operational strategies.

**How do your findings relate to your original question?**

They explain how the two groups differ in terms of trip duration,
timing, and purpose.

**Who is your audience? What is the best way to communicate with them?**

**Audience:** Cyclistic’s executive and marketing teams.

-Use simple and impact visualizations that clearly illustrate
differences (for example, bar charts, line graphs).

-Summarize insights with key statistics (for example, average trip
duration, percentage of rides on weekends).

**Can data visualization help you share your findings?**

Absolutely. Data visualizations make the differences between member and
casual rider behaviors easy to understand. I used the following
visualizations:

✔ Bar chart ✔ Line graph ✔ Stacked bar chart

**Is your presentation accessible to your audience?**

Yes, i am using Labels, legends, titles, highlited colors and
alternative text description for visuals.

#### Key Tasks

**Determine the best way to share your findings.**✔

**Create effective data visualizations.**✔

**Present your findings.**✔

**Make sure your work is accessible.**✔

### Deliverable 5✔ {style="Color: purple"}

**Supporting visualizations and key findings**

Member and Casual Riders Behaviors based on Avg duration, Avg distance
and type of bike.

```{r Comparison table}
print(aggregated_data_member_casual)
```

We can notice a similar bike type usage between casual and members.

```{r}
ggplot(aggregated_long, aes(x = member_casual, y = count, fill = bike_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Comparison of Bike Usage by Member and Casual Riders (Graph 1)",
    x = "Rider Type",
    y = "Number of Rides",
    fill = "Bike Type"
  ) +
  theme_minimal()
```

Similar distance, lower duration (almost 50% lower) for members.

```{r}
ggplot(aggregated_data_duration_distance_long, aes(x = member_casual, y = value, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +  # Create the bar for both metrics
  facet_wrap(~metric, scales = "free_y") +  # Facet metrics
  labs(
    title = "Average Trip Duration and Distance by Member Type (Graph 2)",
    x = "Rider Type",
    y = "Values"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("casual" = "blue", "member" = "green"))
```

Casual riders prefer bikes for recreational activities (Weekends spikes)
and Members for commuting purpose (Weekdays Spikes)

```{r}
ggplot(daily_usage, aes(x = day_of_week, y = total_rides, color = member_casual, group = member_casual)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Cyclistic Bike Usage by Day of the Week (Graph 3)",
    x = "Day of the Week",
    y = "Number of Rides",
    color = "Rider Type"
  ) +
  theme_minimal()
```

## Act

#### Guiding Questions

**What is your final conclusion based on your analysis?**

The analysis reveals that while annual members and casual riders share
similar preferences in bike types and distances traveled, their usage
patterns are distinctly different. Members tend to use the service for
shorter, more frequent trips, likely for commuting purposes during
weekdays. In contrast, casual riders favor longer trips primarily on
weekends, suggesting leisure-oriented use. This distinction underscores
the potential for targeted strategies to convert casual riders into
members by emphasizing membership benefits that cater to their unique
usage patterns.

**How could your team and company apply your insights?**

The insights can be applied in several ways:

**Targeted Marketing:** Launch personalized campaigns that emphasize the
benefits of membership for casual riders, focusing on cost savings for
frequent weekend use, exclusive perks, and flexibility tailored to
leisure activities.

**Incentivized Trial Memberships:** Offer limited-time trial memberships
at a reduced rate or even free for a short period. This allows casual
riders to experience the benefits of being a member without a long-term
commitment, increasing the likelihood of conversion.

**Weekend Membership Plans:** Introduce a special weekend-focused
membership option with lower fees but benefits designed for leisure
users. This strategy caters directly to casual riders who primarily use
the service on weekends.(Just in case option)

**Retention Strategies:** Implement loyalty programs and exclusive
benefits for members, such as discounts on extended rides or priority
access to bikes during peak times, ensuring they see continued value in
maintaining their membership.This aimed at both current members and
newly converted casual riders.

**What next steps would you or your stakeholders take based on your
findings?**

**Evaluate the Feasibility of Proposed Strategies:** Conduct a detailed
analysis of the targeted marketing campaigns, trial memberships, and
weekend-focused plans to assess their viability. This includes
evaluating the potential costs, required resources, and the expected
return on investment (ROI).

**Cost-Benefit Analysis:** Calculate the projected costs for
implementing each strategy, such as marketing expenses, discounts, and
administrative costs. Compare these costs against the anticipated
benefits, including increased membership conversions, higher retention
rates, and overall revenue growth.

**Develop an Implementation Timeline:** Create a step-by-step timeline
for rolling out the proposed strategies. This should include short-term
actions like launching trial memberships and longer-term initiatives
such as introducing new membership plans. Set clear milestones and
deadlines for each phase.

**Monitor and Adjust:** Once implemented, continuously track the
performance of each strategy. Use key performance indicators (KPIs) such
as membership conversion rates, usage patterns, and customer feedback to
assess effectiveness. Be prepared to adjust strategies based on
real-time data to optimize results.

**Is there additional data you could use to expand on your findings?**

**Customer Satisfaction and Feedback Surveys:** Conduct multiple-choice
surveys to gather structured feedback from both casual riders and
members. This will help identify specific areas where the company excels
or needs improvement, providing actionable insights to enhance customer
experience. Understanding what casual riders value most can inform the
design of targeted membership benefits.

**User Journey Data:** Collect data on the complete user journey, from
app interactions to booking and ride completion. Analyzing these
touchpoints can reveal pain points or friction in the user experience,
allowing for targeted improvements that enhance convenience and
satisfaction.

### Deliverable 6✔ {style="Color: purple"}

**My top three recommendations based on my analysis**

#### 1. Targeted Marketing Campaigns

Launch personalized marketing campaigns focused on converting casual
riders into members by highlighting.

✔**Cost savings** for frequent weekend use.

✔**Exclusive perks** tailored to leisure activities.

✔**Flexibility** in membership options, especially for weekend users.
These campaigns should emphasize the benefits casual riders can gain
from membership, addressing their specific usage patterns and
encouraging them to see the value in becoming regular members.

#### 2. Incentivized Trial Memberships

Introduce limited-time trial memberships at a reduced rate or free for a
short period. This allows casual riders to:

✔**Experience the benefits** of membership without long-term commitment.

✔Increase the **likelihood** of conversion by showcasing the convenience
and perks of being a member. This strategy effectively lowers the
barrier to entry, giving casual riders a risk-free opportunity to
explore the advantages of membership.

#### 3. Customer Satisfaction and Feedback Mechanism

Implement structured, multiple-choice customer satisfaction surveys to:

✔Gather **insights on casual riders' preferences** and pain points.

✔Identify areas where the company excels or needs improvement, ensuring
that enhancements are **data-driven**.

✔Use feedback to refine membership benefits, ensuring they align with
what casual riders value most. By continuously collecting and analyzing
customer feedback, the company can make **informed decisions** to
improve user satisfaction and **retention**.
